\chapter{Exercise 11: Replacing Data}

I'm going to show you an alternative way to insert data which helps with
atomic replacement of rows.  You don't necessarily need this too often, but
it does help if you're having to replace whole records and don't want to
do a more complicated \ident{UPDATE} without resorting to transactions.

In this situation, I want to replace my record with another guy but
keep the unique id.  Problem is I'd have to either do a DELETE/INSERT
in a transaction to make it atomic, or I'd need to do a full UPDATE.

Another simpler way to do it is to use the \ident{REPLACE} command, or
add it as a modifier to \ident{INSERT}.  Here's some SQL where I first
fail to insert the new record, then I use these two forms of \ident{REPLACE}
to do it:

\begin{code}{ex11.sql}
<< d['code/ex11.sql|pyg|l'] >>
\end{code}


\section{What You Should See}

In this exercise I'm going to enter these commands in the \program{sqlite3} 
console rather than run them from the command line:

\begin{code}{ex11 SQLite3 Session}
<< d['code/ex11.sqlite3-console|pyg|l'] >>
\end{code}

You can see on line 4 that I get an "Error: PRIMARY KEY must be unique"
because the record has the id 0.  I want to do a combination DELETE/INSERT
with the new record, so then I do an \ident{INSERT OR REPLACE} next.
After that I put the record back with \ident{REPLACE} which is shorthand for
\ident{INSERT OR REPLACE}.

\section{Extra Credit}

\begin{enumerate}
\item Go to the \href{http://www.sqlite.org/lang_insert.html}{SQLite3 INSERT page} and look at the other \ident{INSERT OR} options you have.
\item Practice \ident{REPLACE} by replacing the Unicorn with a pet Parrot.
\end{enumerate}

\section{Portability Notes}

Some databases might not have the \ident{REPLACE} command or even the 
\ident{INSERT OR REPLACE} syntax.  In that case, you'll just have to wait
until I show you transactions.
