\chapter*{Introduction: Haters Gonna Hate, Or Why You Still Need SQL}

SQL无处不在。并不是因为我想要你用它才这么说，而是，这就是一个事实。甚至你的口袋里就有，所有的安卓手机和苹果手机都可以轻易接入一个叫做SQLite的SQL数据库，很多应用程序都是直接调用它的。 很多地方都会使用到SQL：银行，医院，大学，政府，大小型企业。 总之几乎每台电脑每个人都要接触到些跟SQL有关的东西。 SQL是一个相当成功而且靠谱的技术。 

可是呢，SQL的麻烦是， \emph{每个人} 都讨厌它的guts.  SQL是一种古怪的非语言( "non-language" )，这一点大多数程序员都无法忍受。在这些现代的基于网页或对象亲和的编程存在之前，SQL就被设计出来了。 尽管它的操作设计是以坚实的数学理论为基础的， 但它还有足够的麻烦能让你觉得恼人。 树？网化对象？子母层关系？SQL一定会嘲笑你并给你一个混乱的大平面表格说“兄弟诶，自己想吧。”

SQL这么讨厌，你为什么还应该学它呢？那是因为在这讨厌的背后，大多数人都不了解SQL也不知道应该怎样正确的使用它。  NoSQL 运动不仅是对古董数据库服务器的反应的一部分，也是一种害怕SQL而又不知道它到底是如何工作的的一种回应。(The NoSQL movement is partially a reaction to antiquated database servers, and also a response to a fear of SQL borne from ignorance of how it works.) 通过学习SQL，你将学习一些很重要的理论基础，这些理论基础是可以应用到几乎每一种数据存储系统的。

不管SQL反对者如何声称，你都应该学习SQL，要知道SQL无处不在，而且了解它也并没有你想象的那么难。成为一名合格的SQL使用者将在以后帮助你决定使用哪种数据库，是否使用SQL，并从程序员视角给出其他数据系统的深入画面。(Becoming an educated SQL user will help you make informed decisions about what databases to use, whether to not use SQL, and give you a deeper understanding of many of the systems you work with as a programmer.)

最后的最后，我觉得你应该学SQL，因为它很方便。你可以快速的使用 SQLite 来原型开发一个简单的数据模型 (网页应用，手机应用，或者桌面应用), 并且你也知道它几乎可以在拥有 SQLite 的任何平台上使用，这几乎就是哪都能用了。为了拥有这样的跨平台操作能力，也为了使用这么强大的数据存储语言，学习SQL还是很值得滴。

\subsection*{关于本书}

This book teaches SQL to anyone, but it helps you can code already.  The 
concepts in SQL are taught assuming you at least know how to do some programming
even if it's just a tiny bit.  It also assumes you can run commands from the
command line, know how to use the shell, and have access to a good programmer's
text editor.

To keep the book simple, and since managing a giant database server is tangential
to learning SQL, this book will use \href{http://sqlite.org}{SQLite3} to teach
the fundamentals of the language.  It will use SQLite3 similar to how you use
Python or Ruby and you'll be writing full \file{.sql} scripts and running them
to learn the language.  

Each exercise will introduce the concept in the usual "Learn The Hard Way" style.
It will also have a small section on "portability" that explains what might be
different in PostgreSQL and/or MySQL, or any other gotchas to be aware of so that
the book is useful on more than just sqlite3.

When you're done with this book, you should understand SQL well, how to design
data, a bit about optimizing databases, and you should be able to branch out into
another database of your choice.

\subsection*{What Is SQL?}

I pronounce SQL "sequel" but you can also say "ESS-QUEUE-ELL" if you want. SQL
also stands for Structured Query Language but by now nobody even cares about
that since that was just a marketing ploy anyway. What SQL does is give you a
language for interacting with data in a database.  It's advantage though is
that it closely matches a theory established many years ago defining properties
of well structured data.  It's not exactly the same (which some detractors
lament) but it's close enough to be useful.

How SQL works is it understands fields that are in tables, and how to find the
data in the tables based on the contents of the fields.  All SQL operations are
then one of four general things you do to tables:

\begin{description}
\item[Create] Putting data into tables.
\item[Read] Query data out of a table.
\item[Update] Change data already in a table.
\item[Delete] Remove data from the table.
\end{description}

This has been given the acronym "CRUD" and is considered a fundamental set of
features \emph{every} data storage system must have.  In fact, if you can't do 
one of these four in some way then there better be a very good reason.

One way I like to explain how SQL works is by comparing it to a spreadsheet 
software like Excel:

\begin{enumerate}
\item A database is a whole spreadsheet file.
\item A table is a tab/sheet in the spreadsheet, with each one being given a name.
\item A column is a column in both.
\item A row is a row in both.
\item SQL then gives you a language for doing CRUD operations on these to 
    produce \emph{new tables or alter existing ones}.
\end{enumerate}

The last item is significant, and not understanding this causes people a lot of headaches.
SQL only knows tables, and every operation produces tables.  It either "produces" a table
by modifying an existing one, or it returns a new temporary table as your data set.

As you read this book, you'll begin to understand the significance of this design.  For
example, one of the reasons Object Oriented languages are mismatched with SQL databases
is that OOP languages are organized around graphs, but SQL wants to only return tables.
Since it's possible to map nearly any graph to a table and vice-versa this works, but
it places a lot of work on the OOP language to do the translation.  If SQL returned
a nested data structure then this wouldn't be a problem.

Another place that causes a mismatch is in SQL concepts such as ternary
relationships and attributed relationships, which OOP just completely does not
understand.  In SQL I can make 3 tables related to each other using a 4th
table, and that 4th table is a cohesive relationship.  To do the same thing in
an OOP language I have to make a whole intermediary class that encodes this
relationship, which is kind of weird in OOP.

This may sound like total magic incantations right now, but by the time you're
done with this book you'll understand these issues and how to deal with them.

\subsection*{Against Indoctrination}

You may run into someone who thinks you should learn technology X because 
it is superior.  They'll claim that learning SQLite will cripple you for
life because it is missing features.  Or, they may say SQL is dead and
NoSQL is the future.

The problem with these people is they are trying to indoctrinate you, not
educate you.  They think of the world of technology as a zero sum game
that they have to win, and if you learn SQL or SQLite then you won't
learn their system of choice.  Typically, this indoctrination is designed
so that you must depend on their company's software and services to 
stay alive.

I want to educate you so that you have the ability to make your own choices and
learn anything you want.  I'm only using SQLite because it's the simplest and
most complete SQL system you can install.  Other servers are a huge pain to
install and manage, but SQLite is one download, and it's free, and it's
credible. That's the \emph{only} reason I'm using SQLite.  I don't want you to
be stuck on it.  I don't want to convince you to use SQL over NoSQL.

My only goal is to educate you on this particular tool in the simplest way
I can so that you can improve as a programmer, and do it in such a way that
you do \emph{not} need to depend on me or my services when you're done.

Keep that in mind when people try to tell you that learning SQL will somehow
"cripple" you.  Typically they have something to gain by you remaining
ignorant.

\subsection*{License And Status}

This book is in \emph{ALPHA} status right now.  If you have comments on what it
should and should not teach, then feel free to email help@learncodethehardway.org
and let me know.  Commenting on the contents though is pointless since there
really aren't any.

You are free to read this book, but you can't give it out to anyone else or
host it yourself.


